\thetitle

### 2009-01-27 paper/paper.tex
paper: Straightforward fixes suggested by referees.
darcs-hash:20090127222406-bab43-b095ab766e5dc20b9eb16d010f21e8c844b1d67c.gz

\usepackage{color}
\usepackage[colorlinks,pdftitle=\thetitle,pdfauthor=\theauthors]{hyperref}

@@

  Program Fixpoint qs (l: list T) {measure length l}: M (list T) :=
    match l with
    | nil => ret nil
    | pivot :: t =>
        lower <- filterM (gt pivot) t >>= qs;
        upper <- filterM (le pivot) t >>= qs;
        ret (lower ++ pivot :: upper)
    end.
\end{code}
|Monad| is a dependent record containing the (coercible) carrier of type |Set -> Set|, along with |bind| (infix: |>>=|) and |ret| (for ``return'') operations, and the three monad laws. For a general introduction to monadic programming and monad laws, see \cite{wadler93monads}. Furthermore, we use Haskell's ``do-notation'', and freely use standard monadic functions such as |liftM| and |filterM|.

@@

\begin{proof}
  The proof is by induction on |qs|. For |l==nil|, we have {|cost (qs_SP nil) = cost (ret nil) = 0 <= sqrd (length l)|}. For |l==h :: t|, the cost decomposes into
  \begin{code}
    cost (filter (le h) t) + cost (qs_SP (result (filter (lt h) t))) +
    cost (filter (gt h) t) + cost (qs_SP (result (filter (gt h) t))) +
    cost (ret (result (qs_SP (result (filter (lt h) t))) ++
     h :: result (qs_SP (result (filter (gt h) t))))).
  \end{code}
  The |filter| costs are easily proved (by induction on |t|) to be |length t| each. The cost of the final |ret| is 0 by definition. The induction hypothesis applies to the recursive |qs_SP| calls. Furthermore, by induction on |t|, we can easily prove
  \begin{code}
  length (result (filter (le h) t)) + length (result (filter (gt h) t)) <= length t,
  \end{code}
  because the two predicates filtered on are mutually exclusive. Abstracting the filter terms as |flt| and |flt'|, this leaves
  \begin{code}
    length flt + length flt' <= length t ->
      length t + sqrd (length flt) + length t + sqrd (length flt') + 0 <= sqrd (S (length t)),
  \end{code}
  which is true by elementary arithmetic. \qed
\end{proof}

@@

First, we define |partition|. Instead of a two-way comparison operation producing a monadic |bool|, it takes a three-way comparison operation producing a monadic |comparison|, which is an enumeration with values |Lt|, |Eq|, and |Gt|. We represent the resulting partitioning by a function of type |comparison -> list T| rather than a record or tuple type containing three lists, because in the actual source code, this saves us from having to constantly map |comparison| values to corresponding record field accessors or tuple projections.

@@

@@

|expec| gives rise to a host of identities, such as

\begin{code}
  0 <= expec f t
  expec (fun x => f x + g x) t = expec f t + expec g t.
  expec ((* c) fcmp f) = (* c) fcmp expec f

  (forall x `In` t -> f x <= g x) -> expec f t <= expec g t.
  (forall x `In` t -> f x = c) -> expec f t = c
  (forall x, x `In` t -> f x = 0) <-> expec f t = 0

  expec f (t >>= (ret fcmp g)) = expec (f fcmp g) t = expec f (ne_tree.map g t)
\end{code}

@@

  Definition cmp_NDP (x y: T): M bool := ret_ne_tree (1, cmp x y).

@@

  Notation {-"``"-}over m,{-"\ "-} f =O( g ){-"\text{''''}"-} := bigO m f g.
\end{code}
We now state the main theorem.
\begin{code}
  Theorem qs_avg: over length, expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
\end{code}
Thanks to the property discussed at the start of the previous section, % perhaps too far back
|qs_avg| follows as a corollary from the stronger statement that
\begin{code}
  forall l, expec cost (qs_NDP l) <= 2 * length l * (1 + log2ceil (length l)),
\end{code}
the proof of which is described in the next two sections.

@@

After rewriting with the last of the identities listed for |expec| in section \ref{nondetexpec}, the goal becomes
\begin{code}
  expec length (ne_tree.map fst (qs_U li)) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}}"-} ecc i j.
\end{code}
We now invoke another lemma which bounds a nondeterministically computed list's expected length by the expected number of occurrences of specific values in that list. More specifically, it states that

@@

In the interest of brevity, we have omitted lots of gory details and various lemmas in the description of the proof. Still, the parts shown are reasonably faithful to the actual formalization, with two notable exceptions.

>>>>

\usepackage{color}
\usepackage{amsmath}
\usepackage[colorlinks,pdftitle=\thetitle,pdfauthor=\theauthors]{hyperref}

@@

  Program Fixpoint qs (l: list T) {measure length l}: M (list T) :=
    match l with
    | nil => ret nil
    | pivot :: t =>
        lower <- filterM (gt pivot) t >>= qs;
        upper <- filterM (le pivot) t >>= qs;
        ret (lower ++ pivot :: upper)
    end.
\end{code}
|Monad| is a dependent record containing the (coercible) carrier of type |Set -> Set|, along with |bind| (infix: |>>=|) and |ret| (for ``return'') operations, and the three monad laws. For a general introduction to monadic programming and monad laws, see \cite{wadler93monads}. Furthermore, we use Haskell's ``do-notation'', and freely use standard monadic functions such as |liftM| (of type |forall (M: Monad) (A B: Set), (A -> B) -> (M A -> M B)|) and |filterM| (of type |forall (M: Monad) (A: Set), (A -> M bool) -> list A -> M (list A)|).

@@

\begin{proof}
  The proof is by induction on |qs|. For |l==nil|, we have {|cost (qs_SP nil) = cost (ret nil) = 0 <= sqrd (length l)|}. For |l==h :: t|, the cost decomposes into
  \begin{code}
    cost (filter (le h) t) + cost (qs_SP (result (filter (le h) t))) +
    cost (filter (gt h) t) + cost (qs_SP (result (filter (gt h) t))) +
    cost (ret (result (qs_SP (result (filter (le h) t))) ++
     h :: result (qs_SP (result (filter (gt h) t))))).
  \end{code}
  The |filter| costs are easily proved (by induction on |t|) to be |length t| each. The cost of the final |ret| is 0 by definition. The induction hypothesis applies to the recursive |qs_SP| calls. Furthermore, by induction on |t|, we can easily prove
  \begin{code}
  length (result (filter (le h) t)) + length (result (filter (gt h) t)) <= length t,
  \end{code}
  because the two predicates filtered on are mutually exclusive. Abstracting the filter terms as |flt| and |flt'|, this leaves
  \begin{code}
    length flt + length flt' <= length t ->
      length t + sqrd (length flt) + length t + sqrd (length flt') + 0 <= sqrd (S (length t)),
  \end{code}
  which is true by elementary arithmetic. \qed
\end{proof}

@@

First, we define |partition|. Instead of a two-way comparison operation producing a monadic |bool|, it takes a three-way comparison operation producing a monadic |comparison|, which is an enumeration with values |Lt|, |Eq|, and |Gt|. We represent the resulting partitioning by a function of type |comparison -> list T| rather than a record or tuple type containing three lists, because in the actual formalization, this saves us from having to constantly map |comparison| values to corresponding record field accessors or tuple projections. This is only a matter of minor convenience; a record or tuple could have been used instead without problems.

@@

Note that the Quicksort definition in section \ref{embed} could have been implemented with a |partition| pass as well, which might well have made the worst-case proof even simpler. We chose not to do this, in order to emphasize that the properties the average-case proof demands of the algorithm rule out the naive but familiar implementation using |filter| passes.

@@

The function |expec| gives rise to a host of identities, such as

\begin{align}
  0 &\le |expec f t| \nonumber \\
  |expec (fun x => f x + g x) t| &= |expec f t + expec g t| \nonumber \\
  |expec ((* c) fcmp f)| &= |(* c) fcmp expec f| \nonumber \\
  |(forall x `In` t -> f x <= g x) -> expec f t| &\le |expec g t| \nonumber \\
  |(forall x `In` t -> f x = c) -> expec f t| &= c \nonumber \\
  |(forall x `In` t -> f x = 0) <-> expec f t| &= 0 \nonumber \\
  |expec f (t >>= (ret fcmp g))| &= |expec (f fcmp g) t| \nonumber \\
    &= |expec f (ne_tree.map g t)| \label{expec_bind_lem}
\end{align}

@@

  Definition cmp_NDP (x y: T): M_NDP bool := ret_ne_tree (1, cmp x y).

@@

  Notation {-"``"-}over m,{-"\ "-} f =O( g ){-"\text{''''}"-} := bigO m f g.
\end{code}
We now state the main theorem.
\begin{code}
  Theorem qs_avg: over length, expec cost fcmp qs_NDP =O(fun n => INR (n * log2ceil n)).
\end{code}
Thanks to the property discussed at the start of the previous section, % perhaps too far back
|qs_avg| follows as a corollary from the stronger statement
\begin{code}
  qs_expec_cost: forall l, expec cost (qs_NDP l) <= 2 * length l * (1 + log2ceil (length l)),
\end{code}
the proof of which is described in the next two sections.

@@

After rewriting with identity \ref{expec_bind_lem} from section \ref{nondetexpec}, the goal becomes
\begin{code}
  expec length (ne_tree.map fst (qs_U li)) <= {-"\displaystyle{\sum_{(i, j) \in \Varid{IJ}}}"-} ecc i j.
\end{code}
We now invoke another lemma which bounds a nondeterministically computed list's expected length by the expected number of occurrences of specific values in that list. More specifically, it states that

@@

In the interest of brevity, we have omitted lots of detail and various lemmas in the description of the proof. Still, the parts shown are reasonably faithful to the actual formalization, with two notable exceptions.
